---
title: "CoReNat_RSL_3_July_2023"
author: "Rachael Holmes"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

Estimating Holocene relative sea-level for Southeast Asia using the SEAMIS dataset


### First load libraries

```{r, warning = FALSE}

library(tidyverse)

```

### Load data

```{r}
sea <- read.csv("SEAMIS.csv") %>% # seamis datset
  filter(Reject == 0) %>% # removes rejected data
  filter(RSL_m != "n/a")%>% # remove NAs
  mutate(Age_cal_BP = as.numeric(Age_cal_BP))%>%
  filter(Age_cal_BP < 9000) %>% # remove beyond CoReNat data
  # remove data not within focus region
  filter(Sub_region != "Chao Phraya delta") %>% 
  filter(Sub_region != "Song Hong Red River delta") %>% 
  filter(Sub_region!= "Song Hong Red River delta lagoon")%>%
  # kept mekong river delta but not valley as less applicable
  filter(Sub_region!= "Mekong valley")%>% 
  filter(Sub_region!= "Pulu Pandan")%>%
  filter(Region_code != "7") %>%
  # some data with no radiocarbon age still included - remove
  filter(Notes != "No radiocarbon age provided")%>%
  #convert vars to numeric
  mutate(RSL_m = as.numeric(RSL_m))%>%
  mutate(RSL_2sig = as.numeric(RSL_2sig_pos_m))%>%
  mutate(Age_2sig_pos = as.numeric(Age_2sig_pos_cal))%>%
  mutate(Age_2sig_neg = as.numeric(Age_2sig_neg_cal))%>%
  mutate(Region_code = as.factor(Region_code))%>%
  arrange(-Age_cal_BP)%>%
  select(Unique_sample_ID, Reference, Region_code, Sub_region, Latitude, Longitude, Age_cal_BP,Age_2sig_pos, Age_2sig_neg, RSL_m, RSL_2sig)

```

### Plot all data

```{r}

sea %>%
  ggplot(aes(x = Age_cal_BP, y = RSL_m, colour = Sub_region)) +
  geom_point()+
  geom_errorbar(aes(ymin=RSL_m-RSL_2sig, 
                    ymax=RSL_m+RSL_2sig),
                width=.2)+
  geom_errorbar(aes(xmin= Age_2sig_neg, 
                    xmax= Age_2sig_pos),
                width=.2)+
  theme_classic()

```

### Extract predicted RSL from GAM for plotting and using in analysis

```{r, warning = FALSE}

library(mgcv)
library(tidygam)
library(gratia)

# first fit GAM with K = n (obs - 1)

mod_gam <- gam(RSL_m ~ s(Age_cal_BP, k = 112),# long term
                data = sea, method = "REML")# Restricted Maximum Likelihood Estimation to determine wiggliness 

# maybe also need to account for random effect of the site

acf(residuals(mod_gam),main="raw residual ACF")

summary(mod_gam)

gam.check(mod_gam, rep = 500)

appraise(mod_gam, method = "simulate")

plot(mod_gam)

sea %>%
  ggplot(aes(x = Age_cal_BP, y = RSL_m)) +
  geom_point()+
  geom_errorbar(aes(ymin=RSL_m-RSL_2sig, 
                    ymax=RSL_m+RSL_2sig),
                width=.2)+
  geom_errorbar(aes(xmin= Age_2sig_neg, 
                    xmax= Age_2sig_pos),
                width=.2)+
  geom_line(
    colour = "blue", 
    size = 1.2,
    aes(y = fitted(mod_gam))
  ) +
  theme_classic()

```

Pretty good model fit. Explains 78.5% of the variation in the data. K is fairly high, k-index less than 1 but edf and k are not close and p-value is not significant. Allowed maximum amount of k for observations.

Very little autocorrelation in data once filtering SEAMIS more specifically.

Deviance of residuals could be better but generally checks look fine.

```{r}

# predict RSL for every year available in the data 

sea_tidy <- tidygam::predict_gam(mod_gam, length_out = 7905)%>%
  mutate(Age_cal_BP = round(Age_cal_BP))

# make two plots - same dimension etc. to overlay with infinity

t <- c(-69, -50, 0, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 7500, 8000)

# plot of individual points for RSL from SEAMIS

dat <- sea %>%
  ggplot(aes(x= RSL_m, y = Age_cal_BP)) +
  #geom_point(colour = "grey")+
  geom_errorbar(aes(xmin=RSL_m - RSL_2sig, 
                    xmax=RSL_m + RSL_2sig),
                width=.1, colour = "grey")+
  geom_errorbar(aes(ymin= Age_2sig_neg, 
                    ymax= Age_2sig_pos),
                width=.1, colour = "grey")+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10))+
  scale_y_reverse(limits = c(8000, -69), 
                  breaks = c(-69, -50, 0, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 7500, 8000)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  ggtitle("RSL Data")

# plot of predicted RSL based on GAM of SEAMIS data with conf int

gam <- sea_tidy %>%
  ggplot(aes(x= RSL_m, y = Age_cal_BP)) +
  geom_path(colour = "darkorange") +
  geom_ribbon(aes(xmin = lower_ci, xmax = upper_ci),fill = "orange", alpha = 0.1)+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10))+
  scale_y_reverse(limits = c(8000, -69), 
                  breaks = c(-69, -50, 0, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 7500, 8000)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  ggtitle("RSL GAM")

# save predicted RSL for use as env var in ordination

write.csv(sea_tidy, file = "SEA_RSL_gam.csv")

```


### Fit a GAMM with temporal autocorrelation to try to improve model

```{r}

mod_gamm <- gamm(RSL_m ~ s(Age_cal_BP, k = 112),# long term
                correlation = corCAR1(form = ~ Age_cal_BP | Sub_region), data = sea, method = "REML")

acf(residuals(mod_gamm$gam),main="raw residual ACF")

summary(mod_gamm$gam)

gam.check(mod_gamm$gam, rep = 500)

appraise(mod_gamm$gam, method = "simulate")

plot(mod_gamm$gam)

sea %>%
  ggplot(aes(x = Age_cal_BP, y = RSL_m)) +
  geom_point()+
  geom_errorbar(aes(ymin=RSL_m-RSL_2sig, 
                    ymax=RSL_m+RSL_2sig),
                width=.2)+
  geom_errorbar(aes(xmin= Age_2sig_neg, 
                    xmax= Age_2sig_pos),
                width=.2)+
  geom_line(
    colour = "blue", 
    size = 1.2,
    aes(y = fitted(mod_gamm$gam))
  ) +
  theme_classic()


sea_tidy2 <- tidygam::predict_gam(mod_gamm$gam, length_out = 7905)%>%
  mutate(Age_cal_BP = round(Age_cal_BP))

write.csv(sea_tidy, file = "SEA_RSL_gamm.csv")

```
Explains 78.6% of the variation in the data so slightly more than the GAM. K is fairly high, k-index less than 1 but edf and k are not close and p-value is not significant and k-index is higher than for GAM. Allowed maximum amount of k for observations.

Very little change in autocorrelation between GAM and GAMM.

Deviance of residuals could be better but generally checks look fine and an improvement over GAM.

### Plotting interpolated regional RSL from GAMM

```{r}

# plot of predicted RSL based on GAM of SEAMIS data with conf int

gamm <- sea_tidy2 %>%
  ggplot(aes(x= RSL_m, y = Age_cal_BP)) +
  geom_path(colour = "#394F6B") +
  geom_ribbon(aes(xmin = lower_ci, xmax = upper_ci),fill = "#55BEC0", alpha = 0.1)+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10))+
  scale_y_reverse(limits = c(8000, -69), 
                  breaks = c(-69, -50, 0, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 7500, 8000)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  ggtitle("RSL GAMM")

library(ggpubr)

ggarrange(gam, dat, gamm, ncol = 3)


```

### modern changes in SL

```{r}


mod_SL <- read.csv("Ind_mean_change_SL.csv") %>%
  pivot_longer(cols=-year, names_to="satelite", values_to="change_SL")  %>%
  mutate(change_SL_m = change_SL / 1000)%>%
  mutate(Age_BP = 1950 - year )

mod_SL %>%
  ggplot(aes(x = Age_BP, y = change_SL_m, colour = satelite)) +
  geom_point()+
  theme_classic()


mod_SL %>%
  ggplot(aes(x = change_SL_m, y = Age_BP, colour = satelite, shape = satelite)) +
  geom_point()+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10), position = "top")+
  scale_y_reverse(limits = c(8000, -69), 
                  breaks = c(-69, -50, 0, 500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 7500, 8000)) +
  theme_bw() +
  theme(panel.grid = element_blank())+
  scale_colour_brewer(palette = "Set2")
  ggtitle("SL Data")
  
  ## lk1
  
  mod_SL %>%
  ggplot(aes(x = change_SL_m, y = Age_BP, colour = satelite, shape = satelite)) +
  geom_point()+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10), position = "top")+
  scale_y_reverse(limits = c(850, -69), 
                  breaks = c(-69, 70, 100, 150, 200, 250, 
                             300, 350, 400, 450, 500, 550, 
                             600, 650, 700, 750, 800, 850))+
  theme_bw() +
  theme(panel.grid = element_blank())+
  scale_colour_brewer(palette = "Set2")
  ggtitle("SL Data")
  
  #lk2
  mod_SL %>%
  ggplot(aes(x = change_SL_m, y = Age_BP, colour = satelite, shape = satelite)) +
  geom_point()+
  scale_x_continuous(limits = c(-10, 10), breaks = c(-10, 0, 5, 10), position = "top")+
  scale_y_reverse(limits = c(200, -69), 
                  breaks = c(-69, -50, -30, 0, 50, 100, 150, 170, 200))+
  theme_bw() +
  theme(panel.grid = element_blank())+
  scale_colour_brewer(palette = "Set2")

  

```

